% -*- tex-main-file: "main.tex" -*-

\begin{abstract}

\noindent Many sensor network systems expose general interfaces to
system developers for dynamically creating and/or manipulating resources
of various kinds.
%
While these interfaces allow programmers to accomplish common system
tasks simply and efficiently, they also admit the potential for
programmers to mismanage resources.
%
These kinds of errors are particularly problematic for sensor networks,
given the resource constraints and lack of memory protection on current
sensor platforms.



We describe a static analysis that brings the safety of static resource
management to systems that dynamically manage resources.  
%
Our design is driven by the desire for a static, simple, scalable, and
practical analysis framework.
%
Our analysis is based on the observation that sensor network
applications often manipulate resources in a producer-consumer pattern.  
%
In this style, each resource has a unique {\em owner} component at any
given point in time, who has both the sole capability to manipulate the
resource and the responsibility to properly dispose of the resource or
transfer ownership to another component.
%
Our analysis enforces this ownership discipline on components at compile
time.


We have instantiated our approach in Lighthouse, a tool to ensure proper
management of dynamically allocated memory in programs written on top of
the SOS sensor network operating system.
%
We have evaluated the tool on all historical versions of all user
modules in the SOS CVS repository, as well as on the SOS kernel itself.  
%
Our tool generated
%
25 warnings of which 8 were real errors when checking user modules and
35 warnings of which 2 were real errors when checking the kernel,
% 
demonstrating the practical utility of our approach for sensor network
systems.
%
\end{abstract}
