% -*- tex-main-file: "main.tex" -*-

\begin{abstract}
%
\noindent 
Many sensor network systems expose general interfaces to system
developers for dynamically
creating and manipulating resources of various kinds.
%
While these interfaces allow programmers to accomplish
common system tasks simply and efficiently,
they also admit the potential for programmers to mismanage resources,
for example through
leaked resources or improper resource sharing. 
%
We describe a static analysis algorithm and tool that 
brings the safety of static resource management to
systems that dynamically manage resources.
Our analysis is based on the observation that programmers
often use implicit {\em ownership} schemes
to correctly manage dynamic resources.  In such a scheme, each
resource has a unique owner, who has both the capabilities to
manipulate the resource and the responsibilities to use the resource
properly and 
to dispose of it eventually.
Our tool checks resources at compile time for violations of
this ownership discipline.
%

We apply our tool to ensure proper management
of dynamically allocated memory in programs written on top of
SOS, a sensor network operating system.
%
We have evaluated the tool on all historical versions of all user
modules in the SOS CVS repository, as well as on the SOS kernel
itself.  
%
Our tool generated
%
88 warnings of which 16 were real errors when checking user modules and
28 warnings of which 2 were real errors when checking the kernel,
% 
demonstrating the tool's utility for practical sensor network systems.
%
\end{abstract}
