\section{Evaluation}
\label{sec:eval}

In this section we evaluate the need for static checking of resource
management and the practical utility of our
approach for statically detecting resource management errors in
sensor-network software.  Our evaluation uses all code in the
public CVS archive of the SOS operating system.
SOS applications are made up of small user written components called
modules that can be combined into larger programs.  These modules
run on top of the SOS kernel and are developed using an API
supplied by this kernel.


\subsection{Quantifying Dynamic Memory Usage in SOS}

First we measure the prevalence of dynamic resource management within
sensor network systems.  As we have seen, in TinyOS this 
corresponds to the buffer swapping protocol
within messaging stacks, and in SOS
this corresponds to dynamic memory management.
Focusing on SOS, we quantify the usage in SOS modules of the kernel's API for
memory manipulation.
This includes the 
functions used to allocate ({\tt ker\_malloc}), free ({\tt
  ker\_free}), 
and transfer ownership ({\tt SOS\_MSG\_RELEASE} and {\tt
  ker\_msg\_take\_data}) of blocks
of memory.


The SOS CVS head from October 2006 contains 36 modules totaling 5824 source 
lines of code.  Within this sample there are 178 lines of code
calling a kernel function that manipulates memory, or one such function call
for every 33 source lines of code.  A look at all historic version of all
SOS modules reveals that this frequency has not changed significantly
over time.  These findings indicate that dynamic memory management has been
and continues to be an important part of SOS applications.


\subsection{Validating SOS End-User Modules}

Given that
memory is frequently manipulated as a resource in SOS programs, our
next study aimed to answer the following questions:
%
\begin{itemize}
%
\item Do programmers have problems properly managing dynamic memory in SOS?
%
\item Does the analysis and associated checker described in this paper
  help statically identify
those problems?
%
\end{itemize}

To address these questions we ran Lighthouse on 
all end-user modules from SOS
to demonstrate
how it would fit into the ordinary SOS development cycle.  
We applied the checker to every
historic version of each user module included in the SOS CVS
repository that would compile for the \code{mica2} target.  The 48
available modules resulted in a total of 213 unique versions totaling 28042
source lines of code.
%
An external configuration of 26 function annotations was used for this
analysis.
%
Additionally, {\tt lh\_store} annotations were added to 9 module versions,
and a change to the core SOS API from the summer of 2006 required
changing the value of a constant in 6 module versions.


\begin{table}
\caption{Warnings in SOS user modules}
%
\label{tab:module}
\centering 
\begin{tabular}{| l | r |}
    \hline 
    Verified memory leaks identified by analysis & 8 \\
    \hline
    False memory leaks identified by analysis & 8 \\
    \hline 
    Verified dangling pointers identified by analysis & 0 \\
    \hline 
    False dangling pointers identified by analysis & 9 \\
    \hline 
\end{tabular} 
%
\end{table}

Of the 213 modules analyzed, 13 of the modules generated Lighthouse 
warnings for a
total of 25 warnings.
For this
analysis, we only consider the first occurrence of a particular warning
within a module, even if the warning lasts through multiple consecutive
versions of the module.  
%
Each warning was examined by hand and classified as an actual error or
a false positive; the results are presented in Table~\ref{tab:module}.  
%
Eight of the warnings were real memory leaks in the code.
This is impressive since the CVS repository houses well-tested and stable
applications.
%
The 17 remaining warnings were further classified to better understand
the sources of imprecision in Lighthouse.  

\smallskip\noindent{\bf Memory leak false positives.}

False positives from memory leaks came from two different sources of
imprecision in our must-alias analysis.
First, the analysis is intraprocedural, so it must
conservatively
assume that procedure calls can change the contents of pointers passed
as actual arguments.  Therefore, any must-alias information for these
pointers is invalidated across procedure calls, leading to five false
positives.
The analysis can easily rank these as probable false
positives if the end user wishes to see more likely errors first.  The
other three false positives were the result of linked-list data structures
that the must-alias analysis is unable to precisely reason about.  

\smallskip\noindent{\bf Dangling pointer false positives.}

The nine dangling pointer false positives result from limitations of
the may-alias analysis.  
We plan to upgrade this analysis to be both field-sensitive
(distinguishing among the different fields of a structure) and
flow-sensitive (deriving alias information per program point rather
than once for the entire program).  Although these properties can make
an alias analysis difficult to scale in general, sensor network
applications tend to be small due to resource constraints, so we may
be able to leverage these sophisticated techniques in a practical manner.


\subsection{Validating the SOS Kernel}

\begin{table}
\caption{Warnings in the SOS kernel}
%
\label{tab:kernel}
\centering 
\begin{tabular}{| l | r |}
    \hline 
    Verified memory leaks identified by analysis & 2 \\
    \hline
    False memory leaks identified by analysis & 22 \\
    \hline 
    Verified dangling pointers identified by analysis & 0 \\
    \hline 
    False dangling pointers identified by analysis & 11 \\
    \hline 
\end{tabular} 
%
\end{table}


We ran a similar experiment on all code that comprises the current SOS kernel
for the \code{mica2} target.  This configuration consists of 9223
source lines of code and required analyzing 40 source files.  Of these
13 files generated at least one warning for a total of 35 warnings.  A
detailed breakdown of these warnings is provided in
Table~\ref{tab:kernel}.

The two actual errors occurred in
functions that released ownership of some memory to the
caller via a formal parameter,
except 
%
%in exceptional cases where 
%
under specific conditions in which 
%
an error code is returned.
%
The callers of these functions were treating them as unconditionally
releasing memory, without checking the return value.  Lighthouse flags
each function as being in error, since there exist paths to the
function exit that fail to release memory to the formal parameter.  
It would be
interesting to augment our ownership annotations to support the notion
of {\em conditional} ownership transfer.  In such a setting, our
analysis could allow the functions to be considered correct but then
require each caller to check the error code before assuming that it owns
the memory pointed to by the parameter.

While the user modules studied above resulted 
in approximately one false positive for every 1650 source lines of 
code, the SOS kernel had approximately one
false positive for every 270 source lines of code.  
This increase in the rate of false positives is to be expected, since
the kernel includes a significant amount of low-level code that does
not conform to the ownership discipline meant to be obeyed by user
modules.
For example,
the function \code{ker\_msg\_take\_data} is annotated with {\tt
  lh\_claim} to denote that
it returns dynamically allocated memory, but Lighthouse warns that
no such memory is returned.  This is because the function's
implementation directly
accesses the internal data structures of the SOS memory manager to
return a pointer to dynamically allocated data, rather than calling a
function from the SOS API like 
\code{ker\_malloc}.


% ROY (11/04/06):
% This does not seem to stand on its own.  For now I will simply leave
% the allision to this challenge as a point for the conclusion.
%
% \subsection{Comments}
% 
% A third source of false positives arises from application-level
% protocols that ensure memory is used correctly across multiple
% invocations of the module's message handler, but which cannot be
% validated modularly.

\subsection{A Memory Leak in SOS}
\label{ss:tale}

\begin{figure}[tp]
\begin{scriptsize}
\begin{verbatim}
mod_op = (sos_module_op_t*) ker_msg_take_data(msg);
if(mod_op == NULL) return -ENOMEM;
if(mod_op->op == MODULE_OP_INSMOD) {
    existing_module = ker_get_module(mod_op->mod_id);
    if(existing_module != NULL) {
        uint8_t ver = sos_read_header_byte(
                existing_module->header,
                offsetof(mod_header_t, version));
            if (ver < mod_op->version) {
                ker_unload_module(existing_module->pid, 
                        sos_read_header_byte(
                        existing_module->header,
                        offsetof(mod_header_t, version)));
            } else {
                return SOS_OK;
            }
        }
    ret = fetcher_request(KER_DFT_LOADER_PID,
            mod_op->mod_id,
            mod_op->version,
            entohs(mod_op->size),
            msg->saddr);
    s->pend = mod_op;
    ker_led(LED_RED_TOGGLE);
    return SOS_OK;
}
return SOS_OK;
\end{verbatim}
\end{scriptsize}
\caption{\label{fig:leak}A memory leak in an SOS module.}
\end{figure}

This section examines the SOS loader, a core part of SOS that allows
users to dynamically add modules to a running system.  The loader is
known to have caused stability problems for developers in the past.
We did an experiment to understand whether Lighthouse 
could have eased development pains by detecting these problems as they
were introduced into the code.  Since the loader is part of the kernel, 
the historic versions examined in this section were not included in the
analysis of the kernel presented above.

In mid-October 2005 the block of code shown in Figure~\ref{fig:leak}
was checked into CVS as part of {\tt loader.c} and introduced a memory
leak into the loader.  All paths through this block of code leak the
{\tt mod\_op} pointer, which the module acquires ownership of through
{\tt ker\_msg\_take\_data}.  This code results in the following
warning from Lighthouse:

\begin{scriptsize}
\begin{verbatim}
Error: Expression mod_op is not stored after 
        instruction #line 125
mod_op = (sos_module_op_t *)
        ker_msg_take_data((unsigned char)18, msg);
\end{verbatim}
\end{scriptsize}

After three additional revisions that did not significantly modify or
fix the memory leak, a fourth revision was made in mid-December 2005.
This revision expanded the functionality of {\tt loader.c} and broke
the code up into smaller functions:

\begin{scriptsize}
\begin{verbatim}
sos_module_op_t *mod_op;
if (msg->saddr == ker_id() || s->pend) {
    return SOS_OK;
}
mod_op = (sos_module_op_t*) ker_msg_take_data(msg);
if(mod_op == NULL) return -ENOMEM;
switch(mod_op->op){
case MODULE_OP_INSMOD:
    return module_op_insmod(s,msg,mod_op);
case MODULE_OP_RMMOD:
    return module_op_rmmod(s,msg,mod_op);
}
return SOS_OK;
\end{verbatim}
\end{scriptsize}

This code again causes our checker to warn about the memory leak:

\begin{scriptsize}
\begin{verbatim}
Error: Expression mod_op is not stored after 
        instruction #line 186
mod_op = (sos_module_op_t *)
        ker_msg_take_data((unsigned char)18, msg);
\end{verbatim}
\end{scriptsize}

Clearly {\tt mod\_op} is still being leaked if {\tt mod\_op->op} does
not find a matching case in the {\tt switch} statement.  Further,
adding the {\tt lh\_release} attribute to the third formal parameter
of the functions {\tt module\_op\_insmod} and {\tt module\_op\_rmmod}
and re-running the checker reveals that both of these functions also
leak {\tt mod\_op}.

A day later, eight weeks after the memory leaks were first introduced,
these memory leaks were found and fixed:

\begin{scriptsize}
\begin{verbatim}
sos_module_op_t *mod_op;
if (msg->saddr == ker_id() || s->pend) {
    return SOS_OK;
}
mod_op = (sos_module_op_t*) ker_msg_take_data(msg);
if(mod_op == NULL) return -ENOMEM;
switch(mod_op->op){
case MODULE_OP_INSMOD:
    return module_op_insmod(s,msg,mod_op);
case MODULE_OP_RMMOD:
    return module_op_rmmod(s,msg,mod_op);
}
ker_free(mod_op);
return SOS_OK;
\end{verbatim}
\end{scriptsize}

As shown above, a call to {\tt ker\_free} has been added before the
final {\tt return}, in order to properly dispose of {\tt mod\_op}.
The functions {\tt module\_op\_insmod} and {\tt module\_op\_rmmod}
were similarly updated to free their third argument.  This corrected
code passes through the checker without any warnings.  The CVS log
message simply reads, ``Fixed another memory leak.''  



